Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-14 write-ups/PLDI/paper.tex
Shortening.

As with a regular OpenCL kernel (see \mysec\ref{sec:opencl}), a
cooperative kernel is launched by the host application.  The host
application passes parameters to the kernel and specifies a desired
number of workgroups, each consisting of a specified number of
threads.  Unlike in a regular kernel, the parameters to a cooperative kernel are \emph{immutable}, as if
they were qualified with the \texttt{const} keyword (though pointer
parameters can refer to mutable data).

A cooperative kernel is written in OpenCL C plus the following
language extensions: $\transmit$, a
new qualifier on the local variables of a thread; $\offerkill$ and
$\offerfork$, the key functions that enable cooperative scheduling;
$\globalbarrier$ and $\resizingglobalbarrier$, two primitives to allow
inter-workgroup synchronisation.

We now explain the semantics of these language extensions.  We present the semantics intuitively, using English.
In Appendix~\ref{appendix:semantics} (see supplementary material) we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.

@@

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  By
annotating a variable with $\transmit$, the programmer indicates that
when a workgroup spawns new workgroups by calling $\offerfork$, the
workgroup should transmit its current value for the variable to serve
as an initial value for the variable for the threads of the new workgroups.  We detail the semantics
for this when we describe $\offerfork$ below.

@@

The workgroups executing the kernel are called \emph{active
  workgroups}, and $M$ denotes the number of active workgroups.
Initially, the number of active workgroups $M$ is initialised to some
integer between $1$ and $N$---to ensure that at least one workgroup is scheduled, the scheduler will postpone execution of the kernel until a some compute unit becomes available.  The initial active workgroups have
consecutive workgroup ids in the range $[0, M-1]$.

The $\getgroupid()$ function retrieves the id of a workgroup as usual
(see \mysec\ref{sec:opencl}).  When executed by a cooperative
kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups()$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time.

Fair scheduling is guaranteed between active workgroups.  That is, if
some thread in an active workgroup is enabled then eventually some
thread in the active workgroup is guaranteed to execute an
instruction.  Note that our programming model does not mandate that
threads within a workgroup must be fairly scheduled, and indeed this
is often not the case in GPU programming models, because threads are
often organised into \emph{warps} or \emph{wavefronts} within a
workgroup, exhibiting lock-step predicated execution that is
inherently unfair.

@@

The $\offerkill$ primitive allows the cooperative kernel to return
compute units to the scheduler by offering to sacrifice workgroups.
The idea is that while having the scheduler be permitted to terminate
the execution of workgroups in an arbitrary fashion would be drastic,
the kernel may contain identifiable points of execution at which a
workgroup could gracefully leave the computation.  We give examples of
these for work stealing and graph traversal below.

@@

Suppose a workgroup with id $m$ executes $\offerkill$.  
If the workgroup has the highest largest id of all active workgroups then it can be killed by the scheduler, except that workgroup 0 can never be killed (to ensure that the kernel computation does not terminate early).  More formally, if $m < M-1$
or $M=1$, i.e.\ the workgroup's id is not the largest among the active
workgroups, or if only a single workgroup is executing the cooperative
kernel, then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
M-1$, there is a non-deterministic choice between (a) $\offerkill$
executing as a no-op, or (b) execution of the workgroup ceasing and
the number of active workgroups $M$ being atomically decremented by one.  Case
(b) corresponds to the scheduler accepting the workgroup's offer to
cease execution, allowing the scheduler to schedule some other task on
the compute unit on which the workgroup was executing.

>>>>

As with a regular OpenCL kernel (see \mysec\ref{sec:opencl}), a
cooperative kernel is launched by the host application.  The host
application passes parameters to the kernel and specifies a desired
number of workgroups, each consisting of a specified number of
threads.  Unlike in a regular kernel, the parameters to a cooperative kernel are immutable (though pointer
parameters can refer to mutable data).

Cooperative kernels are written using the following 
extensions: $\transmit$, a qualifier on the variables of a
thread; $\offerkill$ and $\offerfork$, the key functions that enable
cooperative scheduling; and $\globalbarrier$ and $\resizingglobalbarrier$
primitives for inter-workgroup synchronisation.
%
We present the semantics of these constructs intuitively, using English.
In Appendix~\ref{appendix:semantics} (see supplementary material) we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.

@@

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  Annotating
a variable $v$ with $\transmit$ means that when a workgroup spawns new
workgroups by calling $\offerfork$, the workgroup should transmit its
current value for $v$ to the threads of the new workgroups, to serve
as an initial value for $v$.  We detail the semantics for this when we
describe $\offerfork$ below.

@@

The number of \emph{active workgroups}---workgroups executing the
kernel---is denoted $M$.  The active workgroups
have consecutive workgroup ids in the range $[0, M-1]$.
Initially, at least one workgroup is active;
if necessary the scheduler must postpone execution of the kernel until
a some compute unit becomes available.

When executed by a cooperative
kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups()$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time (see \mysec\ref{sec:opencl}).

Fair scheduling \emph{is} guaranteed between active workgroups.  That is, if
some thread in an active workgroup is enabled then eventually some
thread in the active workgroup is guaranteed to execute an
instruction.

%Note that our programming model does not mandate that
%threads within a workgroup must be fairly scheduled, and indeed this
%is often not the case in GPU programming models, because threads are
%often organised into \emph{warps} or \emph{wavefronts} within a
%workgroup, exhibiting lock-step predicated execution that is
%inherently unfair.

@@

The $\offerkill$ primitive allows the cooperative kernel to return
compute units to the scheduler by offering to sacrifice workgroups.
The idea is as follows: allowing the scheduler to arbitrarily and abruptly terminate execution
of workgroups might be drastic, yet the kernel
may contain specific program points at which a workgroup could
\emph{gracefully} leave the computation.  We give examples of these
for work stealing and graph traversal in \mysec\ref{sec:programmingguidelines}.

@@

Suppose a workgroup with id $m$ executes $\offerkill$.  
If the workgroup has the highest largest id of all active workgroups then it can be killed by the scheduler, except that workgroup 0 can never be killed (to ensure that the kernel computation does not terminate early).  More formally, if $m < M-1$
or $M=1$ then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
M-1$, the scheduler can choose to ignore the offer, so that $\offerkill$
executes as a no-op, or accept the offer, so that execution of the workgroup ceases and
the number of active workgroups $M$ is atomically decremented by one.  In the second case, the scheduler can assign some other task to
the compute unit on which the workgroup was executing.
